Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-29 paper/paper.tex

\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,}
  $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
\sc{car-clo-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, x,
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
\sc{car-clo}\\
\hline
\end{tabular}
\end{center}
\caption{Minefield semantics for \sc{car-clo-bang} and
  \sc{car-clo} \label{fig:minefield-semantics-for-some}}
\end{figure*}

@@

\subsection{Soundness Results}
We first need an auxiliary  result about minefield semantics. Consider
a trace of  a minefield execution.  For every  minefield state $(\rho,
S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
application $s$) is  non-null.  This can be proved by  an induction on
the number  of steps leading  to the  minefield state.  The  base step
holds   because  the   demand   $\sigma_{all}$   on  $(\mainpgm)$   is
non-null. For the inductive step we  observe that for each step of the
minefield semantics, if the demand $\sigma$  on the LHS of a minefield
step  is non-null,  the  demand  on the  RHS  is  a transformation  of
$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.

@@

%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
%% that the results of liveness analysis of any program can be modeled by
%% the kind  of grammar described above. The  following proposition shows
%% that the converse also holds.
%% \begin{proposition}
%% Given    a     grammar    $G$    of    the     form    described    in
%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
%% program $p$  with program points  $\pi_i$ and variables $x$  such that
%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
%% \end{proposition}
We first show that membership problem of the class of grammars in
Definition~\ref{def:specialgrammar} is undecidable. However note that
the grammars may not be necessarily generated from liveness analysis.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
\begin{lemma}\label{lemma:grammar-undecidable}
Consider a grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only. The membership problem
$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
\end{lemma}

@@

\begin{lemma}\label{lemma:grammar-from-analysis-undecidable}
Consider    a    grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar}  that  is  generated  by  liveness
analysis  of a  program,  and  a forward  access  path $\alpha$.   The
membership problem $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
\end{lemma}

\begin{proof}
Given a Turing machine and an input string, we shall define a function
for   each  of   the  non-terminals   introduced  in   the  proof   of
Lemma~\ref{lemma:grammar-undecidable}.
%% Liveness  analysis  of  $f$  will  result in  not  only  contain  a
%% production  $P$ itself,  but could  also contain  extra productions
%% that necessarily contain the \clazy\ symbol.
For instance, corresponding to the production $\nont{i}{c} \rightarrow
c \nont{j}{\acar}\bcar$, with $c$ as \acar\ for concreteness, consider
the program:

@@

If there more than one productions with the same LHS, then the
corresponding function creates joins them with a $\SIF$.   If there
are $n$ such functions,  then \mainpgm\  creates  a $n$-way
branch  and inserts  a  single call  to a  distinct  function in  each
branch. It can  be verified that the grammar for the function shown
above, includes the productions:
\begin{align*}
\nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{all}{} \\
\nont{i}{\acar} &\rightarrow \clazy\nont{i}{\acar}\bcar\nont{all}{} \\
\nont{all}{} &\rightarrow \epsilon \mid \acar\nont{all}{} \mid
\acdr\nont{all}{}
\end{align*}

Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
had a single non-terminal on the RHS. There are two characteristics of
the grammar produced from liveness-analysis  that are are relevant for
replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
grammar  generated from  liveness  analysis of the constructed program
will  have the  production
$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
\nont{\mathit{all}}{}$,  and (b)  other  productions are  of the  form
$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
\nont{\mathit{all}}{}$, where $\beta_1'$ and $\beta_2'$ do not contain
any non-terminals.

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
replacing   $\nont{\mathit{all}}{}$   in   the  sentential   form   by
$\epsilon$. Conversely,  if the TM  does not accept the  input string,
then  any sentential  form derived  from the  start symbol  would have
exactly one non-terminal, and this non-terminal will be different from
$\nont{\mathit{final}}{c}$.  Thus  the  empty   string  would  not  be
derivable from the grammar.
\end{proof}

@@

In        the       liveness        analysis       described        in
Section~\ref{sec:liveness-analysis}, the liveness of closure variables
in  an application  $s$ is  derived from  dependence chains  along all
paths from the program point where the closure is created to reachable
evaluation  points.  Assume,  for   the  sake  of  concreteness,  that
$\epath_1$  and $\epath_2$  two  such evaluation  points.   This is  a
static approximation  without any  knowledge regarding which  of these
paths will be  eventually taken during execution. During  garbage collection, we would
like to use a more precise  liveness based on the  actual
paths  taken  during  execution.    Therefore we create  liveness  automata
separately  for the dependences in   $\epath_1$ and  $\epath_2$  as  well as  a  automaton
corresponding to  both paths  $\epath_1$ and $\epath_2$.   The closure
corresponding to  $s$ carries  the liveness  environment for  its free
variables (as pointers to automata, one for each variable).  Initially
the  liveness  environment is  based  on  the dependences  along  both
$\epath_1$    and   $\epath_2$.    However,   after    evaluating   an
\SIF\ condition, the liveness environments are updated to one based on
either $\epath_1$ or  $\epath_2$, so that subsequent garbage collections  are based on
more precise liveness information.

@@

\newcommand{\hgt}{2.9cm}
\begin{figure*}[t]
\renewcommand{\arraystretch}{.1}
\begin{tabular}{@{}c@{}@{}c@{}}
%%    \hskip -4mm{\epsfig{file=lambda.eps, height=\hgt}}
%%    & {\epsfig{file=lambda_win.eps, height=\hgt}}
%% \\ \hskip -4mm{\epsfig{file=nperm.eps, height=\hgt}}
%% &  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
%% \\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
%% &  \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
%% \\ \hskip -4mm{\epsfig{file=lcss.eps, height=\hgt}}
%% &  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
%% \\ \hskip -4mm{\epsfig{file=sudoku.eps, height=\hgt}}
%% &  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
%% \\ \hskip -4mm{\epsfig{file=fibheap.eps, height=\hgt}}
%% &  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
%% \\ \hskip -4mm{\epsfig{file=nqueens.eps, height=\hgt}}
%% &  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
%% \\ \hskip -4mm{\epsfig{file=knightstour.eps, height=\hgt}}
%% &  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
   \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
\\   \hskip -4mm{\epsfig{file=treejoin_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
\\   \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
\\  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
%\\ (a) & (b)
\end{tabular}%\vskip -10mm
 \caption{Memory usage.  
%% Column (a) shows complete  usage, column 
%%    (b) enlarges  a part  of the  usage.  
The blue and the red curves indicate the number of cons
cells  in  the  active   semi-space  for  RGC  and  LGC
respectively.  The black curve represents the number of
reachable cells and the  lightblue curve represents the
number  of  cells  that  are actually  live  (of  which
liveness analysis does a static approximation).  x-axis
is the time measured  in number of cons-cells allocated
(scaled down by factor $10^5$). y-axis is the number of
cons-cells (scaled down by $10^3$).}
\label{fig:memory-usage} \figrule
\end{figure*}

>>>>

\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,}
  $\heap(\rho(x)) \mbox{ is } \bot$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc car-clo-bang} 
\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\CAR~x), \sigma) = (\rho, S, \heap)}$,} $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$ & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \ell,
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, \cred{(\clazy \cup \acar)\sigma }$         &
{\sc car-clo}\\
\hline
\end{tabular}
\end{center}
\caption{Minefield semantics for {\sc car-clo-bang} and
  {\sc car-clo} \label{fig:minefield-semantics-for-some}}
\end{figure*}

@@

\subsection{Soundness Results}
We first need an auxiliary  result about minefield semantics. Consider
a trace of  a minefield execution.  For every  minefield state $(\rho,
S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
application $s$) is  non-null.  This can be proved by  an induction on
the number  of steps leading  to the  minefield state.  The  base step
holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
non-null. For the inductive step we  observe that for each step of the
minefield semantics, if the demand $\sigma$  on the LHS of a minefield
step  is non-null,  the  demand  on the  RHS  is  a transformation  of
$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.

@@

%% From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
%% that the results of liveness analysis of any program can be modeled by
%% the kind  of grammar described above. The  following proposition shows
%% that the converse also holds.
%% \begin{proposition}
%% Given    a     grammar    $G$    of    the     form    described    in
%% Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
%% program $p$  with program points  $\pi_i$ and variables $x$  such that
%% the liveness  analysis of $p$  is same as  $G$ except for a  change in
%% non-terminal names which now have the form \var{\Lanv{i}{x}}.
%% \end{proposition}
We first show that membership problem of the class of grammars in
Definition~\ref{def:specialgrammar} is undecidable. However note that
the grammars may not be necessarily generated from liveness analysis.
\newcommand{\state}{\ensuremath{\mathsf{S}}}
\newcommand{\nont}[2]{\ensuremath{\mathsf{S}_{#1}^{#2}}}  
\begin{lemma}\label{lemma:grammar-undecidable}
Given a grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar}  and a forward access path $\alpha$
consisting  of symbols \acar\  and \acdr\  only, the membership problem
$\alpha\$ \in \mathscr{L}(G)$ is undecidable.
\end{lemma}

@@

\begin{lemma}\label{lemma:grammar-from-analysis-undecidable}
Given   a    grammar    $G$    of   the    kind    described    in
Definition~\ref{def:specialgrammar}  that  is  generated  by  liveness
analysis  of a  program  and  a forward  access  path $\alpha$,    the
membership problem $\alpha\$ \in \mathscr{L}(G)$ is undecidable.
\end{lemma}

\begin{proof}
Given a Turing machine and an input string, we shall define a function
for   each  non-terminal   introduced  in   the  proof   of
Lemma~\ref{lemma:grammar-undecidable}.
%% Liveness  analysis  of  $f$  will  result in  not  only  contain  a
%% production  $P$ itself,  but could  also contain  extra productions
%% that necessarily contain the \clazy\ symbol.
For instance, corresponding to the production $\nont{i}{\acar} \rightarrow
\acar \nont{j}{\acar}\bcar$,  consider
the program:

@@

If there more than one productions with the same LHS, then the
corresponding function creates joins them with a $\SIF$.   If there
are $n$ such functions,  then \mainpgm\  creates  a $n$-way
branch  and inserts  a  single call  to a  distinct  function in  each
branch. It can  be verified that the grammar for the function shown
above, includes the productions---$\nont{\mathit{all}}{}$
corresponds to the demand $\sigma_{\mathit{all}}$:
\begin{align*}
\nont{i}{\acar} &\rightarrow \acar\nont{j}{\acar}\bcar\nont{\mathit{all}}{} \\
\nont{i}{\acar} &\rightarrow \clazy\nont{i}{\acar}\bcar\nont{\mathit{all}}{} \\
\nont{\mathit{all}}{} &\rightarrow \epsilon \mid \acar\nont{\mathit{all}}{} \mid
\acdr\nont{\mathit{all}}{}
\end{align*}

Notice that  each production  in Lemma~\ref{lemma:grammar-undecidable}
had a single non-terminal on the RHS. There are two characteristics of
the grammar produced from liveness-analysis  that are are relevant for
replaying the  earlier proof:  (a) If $\mathsf{S}  \rightarrow \beta_1
\mathsf{S'}  \beta_2$  was a  production  in  the earlier  proof,  the
grammar  generated from  liveness  analysis of the constructed program
will  have the  production
$\mathsf{S}      \rightarrow      \beta_1     \mathsf{S'}      \beta_2
\nont{\mathit{\mathit{all}}}{}$,  and (b)  other  productions are  of the  form
$\mathsf{S}       \rightarrow       \beta_1'\mathsf{S'}       \beta_2'
\nont{\mathit{all}}{}$, where $\beta_1'$ and $\beta_2'$ do not contain
any non-terminals.

It is clear that if the TM given as an instance of the halting problem
accepts the input string, then the earlier derivation can be replayed,
every time replacing   $\nont{\mathit{\mathit{all}}}{}$   in   the  sentential   form   by
$\epsilon$. Conversely,  if the TM  does not accept the  input string,
then  any sentential  form derived  from the  start symbol  would have
exactly one non-terminal, and this non-terminal will be different from
$\nont{\mathit{final}}{c}$.  Thus  the  empty   string  would  not  be
derivable from the grammar.
\end{proof}

@@

In        the       liveness        analysis       described        in
Section~\ref{sec:liveness-analysis},   the  liveness   of  a   closure
variable is  derived from dependence  chains along all paths  from the
program point  where the closure  was created to  reachable evaluation
points.  Assume,  for the  sake of  concreteness, that  $\epath_1$ and
$\epath_2$ are two such evaluation points.  During garbage collection,
we would like to use a more precise liveness, based on the actual paths
taken  during  execution.   Therefore   we  create  separate  liveness
automata for dependences  along paths to $\epath_1$  and $\epath_2$, in
addition to  automata for dependences  along paths to  both $\epath_1$
and $\epath_2$.  The closure carries  the liveness environment for its
free  variables (as  pointers  to automata,  one  for each  variable).
Initially the liveness  environment is based on  the dependences along
both  $\epath_1$   and  $\epath_2$.   However,  after   evaluating  an
\SIF\ condition, the liveness environments are updated to one based on
either   $\epath_1$  or   $\epath_2$,  so   that  subsequent   garbage
collections are based on more precise liveness information.

@@

\newcommand{\hgt}{2.9cm}
\begin{figure*}[t]
\renewcommand{\arraystretch}{.1}
\begin{tabular}{@{}c@{}@{}c@{}@{}c@{}}
   \hskip -4mm{\epsfig{file=fibheap_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=sudoku_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nperm_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=paraffins.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lcss_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=huffman_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=knightstour_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=nqueens_win.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=deriv_win.eps, height=\hgt}}
\\ \hskip -4mm{\epsfig{file=treejoin.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=gc_bench.eps, height=\hgt}}
&  \hskip -4mm{\epsfig{file=lambda_win.eps, height=\hgt}}
%\\ (a) & (b)
\end{tabular}%\vskip -10mm
 \caption{Memory usage.  
%% Column (a) shows complete  usage, column 
%%    (b) enlarges  a part  of the  usage.  
The blue and the red curves indicate the number of cons
cells  in  the  active   semi-space  for  RGC  and  LGC
respectively.  The black curve represents the number of
reachable cells and the  lightblue curve represents the
number  of  cells  that  are actually  live  (of  which
liveness analysis does a static approximation).  x-axis
is the time measured  in number of cons-cells allocated
(scaled down by factor $10^5$). y-axis is the number of
cons-cells (scaled down by $10^3$).}
\label{fig:memory-usage} \figrule
\end{figure*}
